---
title: "reslr: advanced"
author: "Maeve Upton, Andrew Parnell and Niamh Cahill"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
       toc: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{reslr: advanced}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
library(tidyverse)
```

## Introduction

In this document we present a range of advanced options that are available in the reslr package. To examine the all other options in the `reslr` package, see the main vignette.

If you have created other examples demonstrating `reslr` capabilities yourself, please [contact me](https://github.com/maeveupton/reslr/issues) so I can include them here for other people to see.

Installing `reslr` and loading it with:
```{r, eval = TRUE}
#install.packages("reslr")
devtools::install_github("maeveupton/reslr")
library(reslr)
```

## Errors in Variables Integrated Gaussian Process with detrended data
The EIV Integrated Gaussian Process model provides the underlying rate of the process directly from the model. If the user requires a model applied to data that has been de-trended this is available in our package for a single site using the `model_type = "eiv_igp_t"`. The data is de-trended using the linear rate which can be provided by the user within the input dataframe along with the uncertainty associated with the rate. Or the linear rate is calculated within the package using the data and this rate is used to de-trend the data. In addition, the user will need to provide the estimated year of core collection.

This technique focuses on 1 site and we do not recommended for multiple proxy sites together. Tide gauge data can be included, however, the user must investigate which tide gauge is suitable and the tide gauge data will use the same linear_rate as the proxy records. As an example, we will filter the example dataset `NAACproxydata` to select one site to demonstrate the process:
```{r,exampledataset2 ,eval = TRUE}
# For 1 site
CedarIslandNC <- reslr::NAACproxydata %>% dplyr::filter(Site == "Cedar Island")
```

**Step 1:** Load in the data using the `reslr_load` function: 
```{r, loadigp ,eval = TRUE}
CedarIslandNC_input_detrend <- reslr_load(
  data = CedarIslandNC,
  include_tide_gauge = FALSE,
  include_linear_rate = TRUE,
  TG_minimum_dist_proxy = FALSE,
  list_preferred_TGs = NULL,
  all_TG_1deg = FALSE,
  prediction_grid_res = 50,
  sediment_average_TG = 10,
  detrend_data = TRUE,
  core_col_year = 2010
)
```
For this case, we update the setting `detrend_data = TRUE` and we provide the `core_col_year = 2010` corresponding to the year of the core collection. The linear_rate is required and is included by setting `include_linear_rate = TRUE`.

The output of this function is a list of two dataframes called `data` and `data_grid`. 
- The `data` dataframe is the inputted data with additional columns for the data_type_id which will contain "ProxyRecord". It can be accessed by:
```{r,dataigp,eval=TRUE}
data <- CedarIslandNC_input_detrend$data
```
- The `data_grid` is a dataframe that is evenly spaced in time based on the `prediction_grid_res` value chosen by the user and is used to create the plots and is accessed by
```{r,datagridigp, eval = TRUE}
data_grid <- CedarIslandNC_input_detrend$data_grid
```

**Step 1a:** A brief insight into the outputs of the `reslr_input` function can be obtained using:
```{r,printigp, eval=TRUE}
print(CedarIslandNC_input_detrend)
```

**Step 2:** Plotting the data the raw data with:
```{r, plotigpdata,fig.align = 'center',fig.width = 7,fig.height = 5,eval = TRUE}
plot(
  x = CedarIslandNC_input_detrend,
  title = "Plot of the raw detrended data",
  xlab = "Year (CE)",
  ylab = "Sea Level (m)",
  plot_proxy_records = TRUE,
  plot_tide_gauges = FALSE
)
```
This will produce a plot of Age on the x-axis and Sea Level on the y-axis in meters. Grey boxes represent the uncertainty associated with the vertical and horizontal uncertainty. The black data points are the midpoints of these uncertainty boxes. The extra arguments can be used which allows the user to updated the titles and axis labels. The caption `plot_caption`, included by default, provides the number of proxy sites and tide gauge sites that will be used in the model and can be removed if required `plot_caption = FALSE`. In addition, the user can select to plot the additional tide gauge data, `plot_tide_gauge = TRUE`.

**Step 3:** To run the eiv_igp_t model the following function should be used:
```{r,runigp, eval = TRUE}
res_eiv_igp_t_detrend <- reslr_mcmc(
  input_data = CedarIslandNC_input_detrend,
  model_type = "eiv_igp_t",
  CI = 0.95
)
```
This command takes the input data and the user specifies the statistical model, i.e. an integrated Gaussian process using the EIV uncertainty method ("eiv_slr_t"). It tells `reslr` to store the output of the model run in an object called `res_eiv_igp_t`. The `CI` setting allows the user to set the credible intervals, the current default is ` CI = 0.95`.

**Step 3a:** A brief insight into the outputs of the `reslr_output` function can be obtained using:
```{r,printigpout, eval=TRUE}
print(res_eiv_igp_t_detrend)
```

**Step 4:** The convergence of the algorithm is examined and he parameter estimates from the model can be investigated using the following:
```{r,summaryigp, eval = TRUE}
summary(res_eiv_igp_t_detrend)
```
If the model run has the package will print: "No convergence issues detected". If the package prints: "Convergence issues detected, a longer run is necessary". The user is required to update the `reslr_mcmc` function with additional iterations in the following manner:
```{r,runigpmore, eval = FALSE}
res_eiv_igp_t_detrend <- reslr_mcmc(
  input_data = CedarIslandNC_input_detrend,
  model_type = "eiv_igp_t",
  # Update these values
  n_iterations = 6000, # Number of iterations
  n_burnin = 1000, # Number of iterations to discard at the beginning
  n_thin = 4, # Reduces number of output samples to save memory and computation time
  n_chains = 3 # Number of Markov chains
)
```

For the parameter estimates, the length scale parameter, "rho" is the correlation parameter and "nu" is the standard deviation of the rate process. "sigma_y" is the variation of the model. 

**Step 5:** The results from the EIV IGP model for the de-trended data can be illustrated using:
```{r,plotigpres, fig.align = 'center',fig.width = 7,fig.height = 5,eval = TRUE}
plot(res_eiv_igp_t_detrend,
  plot_type = "model_fit_plot",
  xlab = "Year (CE)",
  ylab = "Sea Level (m)",
  plot_proxy_records = TRUE,
  plot_tide_gauges = FALSE
)
```
The output of this function is a graph of the input data, i.e. Age and SL and associated uncertainty boxes, and the model fit with 95 \% credible interval. The caption provides the model type used and number of proxy sites and tide gauge sites used and can be removed if necessary with `plot_caption = FALSE`. In order to view the rate of change plot, the following setting should be used:

```{r,plotigpresrate, fig.align = 'center',fig.width = 7,fig.height = 5,eval = TRUE}
plot(res_eiv_igp_t_detrend,
  plot_type = "rate_plot",
  xlab = "Year (CE)",
  y_rate_lab = "Rate of Change (mm per year)"
)
```
This prints the plot of the rate of change with 95 \% credible intervals. The caption provides the model type, the number of proxy sites and tide gauge sites that were used.

To examine the data creating these plots the user types the following:
```{r,igpdataout, eval = TRUE}
output_dataframes <- res_eiv_igp_t_detrend$output_dataframes
head(output_dataframes)
```

## Using a different input Age unit
The default input Age for the reslr package is in units of years in the Common Era. The user can select to input the Age value in Before Present by updating the `input_Age_type = "BP"`. This will alter the output plots as the axis will be reversed. Here is an example of how this is done using the same example site as the previous case but using a Noisy Input spline in time instead.

As an example, we will filter the example dataset `NAACproxydata` to select one site to demonstrate the process:
```{r,exampledataset ,eval = TRUE}
# For 1 site
CedarIslandNC <- reslr::NAACproxydata %>% dplyr::filter(Site == "Cedar Island")
```

**Step 1:** Load in the data using the `reslr_load` function: 
```{r, load ,eval = TRUE}
CedarIslandNC_input_Age_BP <- reslr_load(
  data = CedarIslandNC,
  include_tide_gauge = FALSE,
  include_linear_rate = FALSE,
  TG_minimum_dist_proxy = FALSE,
  list_preferred_TGs = NULL,
  all_TG_1deg = FALSE,
  prediction_grid_res = 50,
  sediment_average_TG = 10,
  input_Age_type = "BP"
)
```
For this case, we update the setting `input_Age_type = "TRUE"BP`. 

The output of this function is a list of two dataframes called `data` and `data_grid`. 
- The `data` dataframe is the inputted data with additional columns for the data_type_id which will contain "ProxyRecord". It can be accessed by:
```{r,data_new,eval=TRUE}
data <- CedarIslandNC_input_Age_BP$data
```
- The `data_grid` is a dataframe that is evenly spaced in time based on the `prediction_grid_res` value chosen by the user and is used to create the plots and is accessed by
```{r,datagrid_new, eval = TRUE}
data_grid <- CedarIslandNC_input_Age_BP$data_grid
```

**Step 1a:** A brief insight into the outputs of the `reslr_input` function can be obtained using:
```{r,print, eval=TRUE}
print(CedarIslandNC_input_Age_BP)
```

**Step 2:** Plotting the data the raw data with:
```{r, plotdata,fig.align = 'center',fig.width = 7,fig.height = 5,eval = TRUE}
plot(
  x = CedarIslandNC_input_Age_BP,
  title = "Plot of the raw data with input Age in Before Present",
  plot_proxy_records = TRUE,
  plot_tide_gauges = FALSE
)
```
This will produce a plot of Age on the x-axis and Sea Level on the y-axis in meters. Grey boxes represent the uncertainty associated with the vertical and horizontal uncertainty. The black data points are the midpoints of these uncertainty boxes. The extra arguments can be used which allows the user to updated the titles and axis labels. The caption `plot_caption`, included by default, provides the number of proxy sites and tide gauge sites that will be used in the model and can be removed if required `plot_caption = FALSE`. In addition, the user can select to plot the additional tide gauge data, `plot_tide_gauge = TRUE`.

**Step 3:** To run the ni_sp_t model the following function should be used:
```{r,runsp, eval = TRUE}
res_ni_sp_t_age_BP <- reslr_mcmc(
  input_data = CedarIslandNC_input_Age_BP,
  model_type = "ni_sp_t",
  CI = 0.95
)
```
This command takes the input data and the user specifies the statistical model, i.e. an integrated Gaussian process using the EIV uncertainty method ("eiv_slr_t"). It tells `reslr` to store the output of the model run in an object called `res_ni_sp_t_age_BP`. The `CI` setting allows the user to set the credible intervals, the current default is ` CI = 0.95`.

**Step 3a:** A brief insight into the outputs of the `reslr_output` function can be obtained using:
```{r,printout, eval=TRUE}
print(res_ni_sp_t_age_BP)
```

**Step 4:** The convergence of the algorithm is examined and he parameter estimates from the model can be investigated using the following:
```{r,summarynew, eval = TRUE}
summary(res_ni_sp_t_age_BP)
```
If the model run has the package will print: "No convergence issues detected". If the package prints: "Convergence issues detected, a longer run is necessary". The user is required to update the `reslr_mcmc` function with additional iterations.

**Step 5:** The results from the EIV IGP model for the de-trended data can be illustrated using:
```{r,plotres_new, fig.align = 'center',fig.width = 7,fig.height = 5,eval = TRUE}
plot(res_ni_sp_t_age_BP,
  plot_type = "model_fit_plot",
  plot_proxy_records = TRUE,
  plot_tide_gauges = FALSE
)
```
The output of this function is a graph of the input data, i.e. Age and RSL and associated uncertainty boxes, and the model fit with 95 \% credible interval. The caption provides the model type used and number of proxy sites and tide gauge sites used and can be removed if necessary with `plot_caption = FALSE`. In order to view the rate of change plot, the following setting should be used:

```{r,plotresrate, fig.align = 'center',fig.width = 7,fig.height = 5,eval = TRUE}
plot(res_ni_sp_t_age_BP,
  plot_type = "rate_plot",
  y_rate_lab = "Rate of Change (mm per year)"
)
```
This prints the plot of the rate of change with 95 \% credible intervals. The caption provides the model type, the number of proxy sites and tide gauge sites that were used.

To examine the data creating these plots the user types the following:
```{r,dataoutnew, eval = TRUE}
output_dataframes <- res_ni_sp_t_age_BP$output_dataframes
head(output_dataframes)
```


## Including Tide Gauge data
In this package, we have described three ways to select a tide gauge and in the example we showed how to include the closest tide gauge to the proxy site. Here we are demonstrating the other options available to the user.
```{r,data2sites, eval = FALSE}
# For 2 site
multi_site <- reslr::NAACproxydata %>%
  dplyr::filter(Site %in% c("Cedar Island", "Nassau"))
```

If the user has a list of chosen tide gauges from PSMSL website, include them in a list in the `reslr_load` function as demonstrated: 
```{r, eval = FALSE}
multi_site <- reslr_load(
  data = multi_site,
  include_tide_gauge = TRUE,
  include_linear_rate = TRUE,
  TG_minimum_dist_proxy = FALSE,
  # There is no limit to the number of tide gauges provided in the list
  list_preferred_TGs = c(
    "ARGENTIA", "MAYPORT",
    "JACKSONVILLE", "LAKE WORTH PIER",
    "MAYPORT (BAR PILOTS DOCK), FLORIDA"
  ),
  all_TG_1deg = FALSE,
  prediction_grid_res = 50,
  sediment_average_TG = 10
)
```
Here is how to plot this example:
```{r,fig.align = 'center',fig.width = 7,fig.height = 5,eval = FALSE}
plot(
  x = multi_site,
  title = "Plot of the raw data",
  xlab = "Year (CE)",
  ylab = "Relative Sea Level (m)",
  plot_tide_gauges = TRUE,
  plot_proxy_records = TRUE,
  plot_caption = TRUE
)
```

If the user requires all tide gauges within 1 degree of the proxy site, which is the equivalent of 111kms, this can be done by updating `all_TG_1deg=TRUE` in the following way:

```{r, eval = FALSE}
multi_site <- reslr_load(
  data = multi_site,
  include_tide_gauge = TRUE,
  include_linear_rate = TRUE,
  TG_minimum_dist_proxy = FALSE,
  list_preferred_TGs = NULL,
  all_TG_1deg = TRUE,
  prediction_grid_res = 50
)
```
Here is how to plot this example:
```{r,fig.align = 'center',fig.width = 7,fig.height = 5,eval = FALSE}
plot(
  x = multi_site,
  title = "Plot of the raw data",
  xlab = "Year (CE)",
  ylab = "Relative Sea Level (m)",
  plot_tide_gauges = TRUE,
  plot_proxy_records = TRUE,
  plot_caption = TRUE
)
```

## Plotting techniques
In the package, all plot labels for results, i.e. x and y labels and titles, can be updated in the following manner:
```{r, eval = FALSE}
# Example
final_plots <- plot(jags_output = reslr_mcmc(CedarIslandNC, model_type = "ni_spline_t"))
final_plots$plot_result
# Adding new title to the total model fit plot
final_plots$plot_result + ggplot2::ggtitle("New Title Added as Example")
final_plots$plot_result + ggplot2::xlab("New x axis label Added as Example")
final_plots$plot_result + ggplot2::ylab("New y axis label Added as Example")
```

## Accessing the posterior samples
In the package the user has the ability access all the posterior sample for each unknown parameter in the following way:
```{r, eval = FALSE}
# Example
CedarIslandNC_input <- reslr_load(
  data = CedarIslandNC)
res_eiv_slr_t <- 
  reslr_mcmc(CedarIslandNC_input, 
             model_type = "eiv_slr_t")
res_eiv_slr_t$noisy_model_run_output$BUGSoutput$sims.list$beta
```
