---
title: "reslr: quick start guide"
author: "Maeve Upton, Andrew Parnell and Niamh Cahill"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
       toc: true
vignette: >
  %\VignetteIndexEntry{reslr: quick start guide}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

## Step 1: install `reslr`

Use:
```{r, eval = FALSE}
# Not on CRAN yet
#install.packages("reslr")
devtools::install_github("maeveupton/reslr")
```
then,
```{r, message=FALSE}
library(reslr)
```

## Step 2: load in the data

Some proxy record data is included with this package. Find where it is with:

```{r, eval = FALSE}
system.file("extdata", "one_data_site.csv", package = "reslr")
```


## Step 3: load the data into `reslr`
```{r, loadslr, eval = FALSE}
CedarIslandNC_input <- reslr_load(
  data = CedarIslandNC,
  include_tide_gauge = FALSE,
  include_linear_rate = FALSE,
  TG_minimum_dist_proxy = FALSE,
  list_preferred_TGs = NULL,
  all_TG_1deg = FALSE,
  prediction_grid_res = 50,
  input_Age_type = "CE",
  sediment_average_TG = 10
)
```

## Step 4: plot the data

```{r, plotdata,fig.align = 'center',fig.width = 7,fig.height = 5,eval = FALSE}
plot(
  x = CedarIslandNC_input,
  title = "Plot of the raw data",
  xlab = "Year (CE)",
  ylab = "Relative Sea Level (m)",
  plot_tide_gauges = FALSE,
  plot_caption = TRUE
)
```

## Step 5: Select your modelling technique

There are a number of modelling options available to the user:

| Statistical Model | Model Information | `model_type` code | Use |
| ---- | ------- | -- |
| Errors in variables simple linear regression | A straight line of best fit taking account of any age and measurement errors in the RSL values using the method of Cahill et al (2015). Use for single proxy site. | **"eiv_slr_t"** |
| Errors in variables change point model | An extension of the linear regression modelling process. It uses piece-wise linear sections and estimates where/when trend changes occur in the data (Cahill et al. 2015). | **"eiv_cp_t"** |
| Errors in variables integrated Gaussian Process | A non linear fit that utilities a Gaussian process prior on the rate of sea-level change that is then integrated (Cahill et al. 2015). | **"eiv_igp_t"** |
| Noisy Input spline in time | A non-linear fit using regression splines using the method of Upton et al (2023). | **"ni_spline_t"** |
| Noisy Input spline in space and time | A non-linear fit for a set of sites across a region using the method of Upton et al (2023). | **"ni_spline_st"**|
| Noisy Input Generalised Additive model for the decomposition of the RSL signal | A non-linear fit for a set of sites across a region and provides a decomposition of the signal into regional, local-linear (commonly GIA) and local non-linear components. Again this full model is as described in Upton et al (2023). | **"ni_gam_decomp"** |

## Step 6: Run your model and check convergence

```{r, runslr,eval = FALSE}
res_eiv_slr_t <- reslr_mcmc(
  input_data = CedarIslandNC_input,
  model_type = "eiv_slr_t",
  CI = 0.95
)
```

The convergence of the algorithm is examined and he parameter estimates from the model can be investigated using the following:
```{r, summaryslr, eval = FALSE}
summary(res_eiv_slr_t)
```

## Step 7: Plot the results
The results from the eiv_slr_t model can be visualised using the following function:
```{r, plotslrres, fig.align = 'center',fig.width = 7,fig.height = 5,eval = FALSE}
plot(res_eiv_slr_t,
  xlab = "Year (CE)",
  ylab = "Relative Sea Level (m)"
)
```
To examine the data creating these plots the user types the following:
```{r, dataframeslrres, eval = FALSE}
output_dataframes <- res_eiv_slr_t$output_dataframes
head(output_dataframes)
```


To examine the additional options in the `reslr` package, see the main vignette.
